NBA Salaries Exploration by Ekaterina Kushnir
========================================================

This report explores a dataset of NBA salaries for the past ten years 
and players performance statistics over regular NBA seasons from 2007/2008
to 2016/2017. The datasets were taken from kaggle.com 
(https://www.kaggle.com/drgilermo/nba-players-stats/home) and from data.world
(https://data.world/datadavis/nba-salaries/). The source for both datasets is 
the website https://www.basketball-reference.com/. The final dataset was merged
from the datasets described above and filtered so that we have statistics on 
players performance and their salaries going back 10 years. 

Players performance statistics includes number of games (G), overall minutes 
played (MP), widely used advanced game statistics such as player efficiency 
rating (PER), true shooting (TS.), statistics on players perfomance in relation 
to the number of overall available opportunities (such as blocks, steals, 
rebounds  percentatges) as well as simple averages per game such as points per
game (PPG), assists per game (ASTPG), total rebounds per game (TRBPG) etc. 
Please refer to this glossary for the full list of basketball performance 
indicators: https://www.basketball-reference.com/about/glossary.html

I filtered the dataset so that only statistics of those players are considered
who made more than 5 games for a given team in a given season in order to remove 
outliers.

Furthermore, salaries data was adjusted for inflation and will be presented in
2018 dollar terms.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages
library(ggplot2)
library(tidyr)
library(dplyr)
library(memisc)
library(MASS)
library(GGally)
library(scales)
library(reshape2)
library(gridExtra)
library(knitr)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Load the Data
# Load data on seasons statistics and player infromation
player_data <- read.csv('player_data.csv')

player_data <- player_data %>% 
  dplyr::select('name', 'year_start', 'year_end') %>% 
  filter(year_start > 2008 - 21) %>% # longest NBA career is about 21 years
  dplyr::rename(Player = name)

players <- read.csv('Players.csv') %>% 
  filter(born > 2008 - 40) # oldest player is Kurt Thomas with 40 years

seasons_stats <- read.csv('Seasons_Stats.csv') %>% 
  filter(Year >= 2008) 

# merge the data sets, filter and drop variables which won't be needed
levpl <- unique(c(levels(seasons_stats$Player), 
                 levels(players$Player)))
seasons_stats$Player <- factor(seasons_stats$Player, levels = levpl)
players$Player <- factor(players$Player, levels = levpl)
seasons_stats <- left_join(players, seasons_stats, by = 'Player')

levpl <- unique(c(levels(seasons_stats$Player), 
                 levels(player_data$Player)))
seasons_stats$Player <- factor(seasons_stats$Player, levels = levpl)
player_data$Player <- factor(player_data$Player, levels = levpl)
seasons_stats <- left_join(seasons_stats, player_data, by = 'Player')

seasons_stats <- seasons_stats %>% 
  mutate(X.x = NULL, X.y = NULL, birth_city = NULL, birth_state = NULL,
         collage = NULL, blanl = NULL, blank2 = NULL)%>% 
  filter(Year > 2008) %>% 
  filter(Age < 41) %>% 
  mutate(career_year = Year - year_start, 
         age_at_car_start = year_start - born) %>% 
    dplyr::rename(team = Tm) %>% 
  filter(age_at_car_start < 40 & Year - year_end < 0 & team != 'TOT')

seasons_stats$team <- dplyr::recode(seasons_stats$team, 
                                    BRK = 'BKN', CHO = 'CHA', NOP = 'NOH')

# Load and adjust salaries data
nba_salary <- read.csv('nba_salaries_1990_to_2018.csv')

nba_salary <- nba_salary %>% 
  dplyr::rename(Player = player, Year = season_start)

nba_salary <- nba_salary %>% 
  mutate(salary = salary/ 10^6) %>% # show salary in mUSD
  filter(Year > 2011) %>% 
  dplyr::select('Player', 'salary', 'Year', 'team')

nba_salary$team <- dplyr::recode(nba_salary$team, NJN = 'BKN')

# Load inflation data, source: IMF inflation rates for the U.S.
# historical rates 2008 - 2017, forecast 2018
infl_rates <- c(3.8, -0.3, 1.6,	3.1,	2.1,	1.5,	1.6,	0.1,	1.3,	2.1,	2.5) 
Year <- seq(2008, 2018, 1)
infl <- data.frame(Year, infl_rates)
infl$factor <- with(infl, 1 + infl_rates/100)

for (i in 1:11) {
    infl$adj[i] <- prod(infl$factor[(i+1):11])
}

# salaries inflated
nba_salary <- left_join(nba_salary, infl, by = 'Year')
nba_salary <- nba_salary %>%  mutate(salary_infl = salary*adj, 
                                     infl_rates = NULL, factor = NULL)

# join seasons stats and salary data
levpl <- unique(c(levels(seasons_stats$Player), 
                 levels(nba_salary$Player)))
seasons_stats$Player <- factor(seasons_stats$Player, levels = levpl)
nba_salary$Player <- factor(nba_salary$Player, levels = levpl)

nba_stats_salar <- left_join(seasons_stats, nba_salary,
                             by = c('Player', 'Year')) %>% 
  filter(!is.na(salary) & G > 5)

nba_stats_salar <- nba_stats_salar %>% 
  dplyr::rename(Tm_curr_s = team.x, Tm_next_s = team.y)

# reorder the columns so they are easier to read  
nba_stats_salar <- nba_stats_salar[, c(1:8, 54:61, 9:53)]
nba_stats_salar <- nba_stats_salar[, c(1:8, 14, 9:13, 15:61)]

# calculate some more statistics per game
nba_stats_salar <- nba_stats_salar %>% 
  mutate(PPG = PTS / G, MPG = MP / G, PFG = PF / G, FTPG = FT / G, 
         X3PPG = X3P / G, X2PPG = X2P / G, TRBPG = TRB / G, ASTPG = AST / G, 
         STLPG = STL / G, BLKPG = BLK / G, TOVPG = TOV / G)

# define whether player stayed with the same team since last season
# set factor levels for both columns, 1  indicates that the player stayed with
# the team and 0 indicates leaved
levtm <- unique(c(levels(nba_stats_salar$Tm_curr_s), 
                 levels(nba_stats_salar$Tm_next_s)))
nba_stats_salar$Tm_curr_s <- factor(nba_stats_salar$Tm_curr_s, levels = levtm)
nba_stats_salar$Tm_next_s <- factor(nba_stats_salar$Tm_next_s, levels = levtm)
nba_stats_salar$stayed <- 
  ifelse(nba_stats_salar$Tm_curr_s == nba_stats_salar$Tm_next_s, TRUE, FALSE)

# set factor levels for Pos (position)
nba_stats_salar$Pos <- factor(nba_stats_salar$Pos)

```

# Univariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_1}
# print summaries
dim(nba_stats_salar)
str(nba_stats_salar)
summary(nba_stats_salar)

```

The dataset consists of more than 2,000 observation of 73 variables. We will not
be exploring all of them however. Some of them were calculated based on others, 
e.g. field goal percentage (FG.) is calculated using field goals (FG) and field 
goals attemps (FGA). We will be concentrating on commonly used easy, to 
understand basketball statistics and additionally have a closer look at some 
advanced performance statistics. We will try to avoid using variables which are 
a product of other variables. Variables which we will be exploring are salary 
adjusted for inflation, career year, player's age, number of games played for 
the team in a given season, various statistics per game: points, minutes played,
two and three point field goals, free throws, assists, steals, rebounds, 
turnovers, blocks. Furthermore, true shooting percentage which represents the
measure of shooting efficiency and is calculated as a ratio of goals including 
free throws to attempts made will be explored. Additionally, we will have a look
at player effeciency rating (PER). Finally, we will look at whether players 
stayed or leaved the team during or after the season. Since the performance 
statistics mentioned above are influenced by the position the player, we may 
increase our understanding of data by looking at statistics for different 
positions separately later on in the course of the analysis.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_2}
# salary histogram
s1 <- ggplot(aes(x = salary_infl), data = nba_stats_salar) + 
  geom_histogram(binwidth = 1) + scale_x_continuous(breaks = c(1, 10, 20, 30)) + 
  ggtitle('Salary histogram')

# salary log transformed histogram
s2 <- ggplot(aes(x = salary_infl), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) + 
  scale_x_continuous(trans = log10_trans()) + 
  ggtitle('Salary histogram log transformed')

# salary sqrt transformed histogram
s3 <- ggplot(aes(x = salary_infl), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) + 
  scale_x_continuous(trans = sqrt_trans(), 
                     breaks = c(0.1, 0.5, 1, 10, 20, 30)) +
  ggtitle('Salary histogram sqrt transformed')

grid.arrange(s1, s2, s3)

```

Transforming the salary using the square root function allows a more closer look
at the data in the long tail and appears to result in a bimodal distribution 
with peaks at around 0.1 million USD and just above 1 million USD. Let's have a 
look at age and career year of the players.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_PLots_3}
# Age
ggplot(aes(x = Age), data = nba_stats_salar) + geom_histogram(binwidth = 1) +
  scale_x_continuous() + 
  ggtitle('Age')

```

Players age seems to peak at 23-24 years and starts to decrease from there. 
There is only a small portion of players aged over 35.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_PLots_4}
#career year
ggplot(aes(x = career_year), data = nba_stats_salar) + 
  geom_histogram(binwidth = 1) +
  ggtitle('career year')

```

Also most players also appear to be in their early years of NBA career with the 
largest number of players in their first career year. Number of players in later
career years gradually decreases. Only a small portion of players achieve a long
career over 15 years. Let's have a look at performance indicators distributions
(average statistics per game) next.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6, Univariate_PLots_5}
# Simple per game statistics - points
# Points per game 
ppg1 <- ggplot(aes(x = PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 1) +
  scale_x_continuous() + ggtitle('Points per game')
 
ppg2 <- ggplot(aes(x = PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous(trans = sqrt_trans()) + ggtitle('Points per game sqrt')

ppg3 <- ggplot(aes(x = PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous(trans = log_trans()) + ggtitle('Points per game sqrt')

# 2 points goals per game 
x2ppg1 <- ggplot(aes(x = X2PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous() + ggtitle('2 points goals per game')

x2ppg2 <- ggplot(aes(x = X2PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('2 points goals per game sqrt')

x2ppg3 <- ggplot(aes(x = X2PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('2 points goals per game log')

# 3 points goals per game
x3ppg1 <- ggplot(aes(x = X3PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous() + ggtitle('3 points goals per game')

x3ppg2 <- ggplot(aes(x = X3PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('3 points goals per game sqrt')

x3ppg3 <- ggplot(aes(x = X3PPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('3 points goals per game log')

# Free throws goals per game 
ftpg1 <- ggplot(aes(x = FTPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous() + ggtitle('Free throws per game')

ftpg2 <- ggplot(aes(x = FTPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('Free throws per game sqrt')

ftpg3 <- ggplot(aes(x = FTPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('Free throws per game log')

grid.arrange(ppg1, x2ppg1, x3ppg1, ftpg1, 
             ppg2, x2ppg2, x3ppg2, ftpg2, 
             ppg3, x2ppg3, x3ppg3, ftpg3, ncol = 4)

```

As we can see from the plots of points related statistics, transforming the 
scale by the square root and the log functions bring the best view of long 
tails. Using the square root function with the variables points per game, 2 
points per game per game delivers a bell shape curve which looks like normal 
distribution. For the variable 3 points per game we have a surprisingly large 
number of players who haven't scored a single 3 points goal. For the variable 
free throws per game the distribution is right skewed with some outliers scoring
much more free throws than the average. 

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=12, fig.height=6, Univariate_PLots_6}
# Assists per game
astpg1 <- ggplot(aes(x = ASTPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous() + 
  ggtitle('Assists per game')

astpg2 <- ggplot(aes(x = ASTPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('Assists per game sqrt')

astpg3 <- ggplot(aes(x = ASTPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('Assists per game log')

# Steals per game
stlpg1 <- ggplot(aes(x = STLPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous() + 
  ggtitle('Steals per game')

stlpg2 <- ggplot(aes(x = STLPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.05) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('Steals per game sqrt')

stlpg3 <- ggplot(aes(x = STLPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('Steals per game log')

# Total rebounds per game
trbpg1 <- ggplot(aes(x = TRBPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous() + 
  ggtitle('Total rebounds per game')

trbpg2 <- ggplot(aes(x = TRBPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('Total rebounds per game sqrt')

trbpg3 <- ggplot(aes(x = TRBPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('Total rebounds per game log')

# Blocks per game
blkpg1 <- ggplot(aes(x = BLKPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous() + 
  ggtitle('Blocks per game')

blkpg2 <- ggplot(aes(x = BLKPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.05) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('Blocks per game sqrt')

blkpg3 <- ggplot(aes(x = BLKPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('Blocks per game log')

grid.arrange(astpg1, stlpg1, trbpg1, blkpg1, 
             astpg2, stlpg2, trbpg2, blkpg2,
             astpg3, stlpg3, trbpg3, blkpg3,
             ncol=4)

```

By plotting further important basketball statistics such as assists, steals, 
total rebounds and blocks we see a similar picture as for the scoring 
statistics: transforming the scale using the square root and the log function 
helps to understand longer right tales. With blocks we see a similar story as 
with 3 point goals: there is a large number of players who blocked no shots. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_PLots_7}

# Turnovers per game
tovg1 <- ggplot(aes(x = TOVPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous() + 
  ggtitle('Turnovers per game')

tovg2 <- ggplot(aes(x = TOVPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('Turnovers per game sqrt')

tovg3 <- ggplot(aes(x = TOVPG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('Turnovers per game log')


# Personal fouls per game
pfpg1 <- ggplot(aes(x = PFG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) +
  scale_x_continuous() + 
  ggtitle('Personal fouls per game')

pfpg2 <- ggplot(aes(x = PFG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = sqrt_trans()) + 
  ggtitle('Personal fouls per game sqrt')

pfpg3 <- ggplot(aes(x = PFG), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.1) +
  scale_x_continuous(trans = log_trans()) + 
  ggtitle('Personal fouls per game log')

grid.arrange(pfpg1, tovg1, pfpg2, tovg2, pfpg3, tovg3, ncol = 2)

```

Lastly, there are plots of negative performance indicators, personal fouls and
turnovers. Personal fouls per game distribution already has a bell curve shape 
without transforming the scale. Turnovers which indicate loosing the ball are
distributed in a bell curve shape after transforming the scale by using a square 
root function.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_PLots_8}
# Advanced statistic
# Effective field goal percentage
efg <- ggplot(aes(x = eFG.), data = nba_stats_salar) +
  geom_histogram(binwidth = 0.01) + 
  scale_x_continuous()

# True shooting percentage
ts <- ggplot(aes(x = TS.), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.01) + 
  scale_x_continuous()

grid.arrange(efg, ts)

```  

As for effective field goal and true shooting they both look like a bell curve
with high kurtosis with the most common values about 0.5 for effective field
goal and slightly above 0.5 for true shooting. This makes sense since effective
field goal percentage takes into account shots accuracy using shots from the
field only (both 2 and 3 points shots) and true shooting additionally considers 
shots accuracy of free throws. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_PLots_9}
# Player efficiency rating 
ggplot(aes(x = PER), data = nba_stats_salar) + geom_histogram(binwidth = 1) + 
  scale_x_continuous(breaks = c(-20, 0, 10, 15, 20, 30, 40))

```

Finally, from the player efficiency rating distribution plot we can conclude 
that the most common PER level lies between 10 and 15 and only few player have 
achieved a PER above 20. PER is widely used indicator of player's performance. 
PER is calculated using a specially developed formula taking into account a wide
range of performance indicators of an individuall player adjusted by specially 
developed factors. Please refer to 
https://www.basketball-reference.com/about/per.html for further details.

Let's have a look at salary distribution of player with PER below 15 as well as
above 20 together with the salary histogram of all players.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_10}
# salary histogram
s_all <- ggplot(aes(x = salary_infl), data = nba_stats_salar) + 
  geom_histogram(binwidth = 1) + 
  scale_x_continuous(breaks = c(0, 10, 20, 30), limits = c(0, 40)) + 
  ggtitle('Salary histogram - all')

# Player efficiency rating below or equal 20 
s_low <- ggplot(aes(x = salary_infl), 
                data = subset(nba_stats_salar, PER <= 20)) + 
  geom_histogram(binwidth = 0.5) + 
  scale_x_continuous(breaks = c(0, 10, 20, 30), 
                     limits = c(0, 40)) + 
  ggtitle('Salary histogram with PER at or below 20')

# Player efficiency rating above 20
s_high <- ggplot(aes(x = salary_infl), 
                 data = subset(nba_stats_salar, PER > 20)) + 
  geom_histogram(binwidth = 1.5) + 
  scale_x_continuous(breaks = c(0, 10, 20, 30), 
                     limits = c(0, 40)) + 
  ggtitle('Salary histogram with PER over 20')

grid.arrange(s_all, s_low, s_high)

```

The salaries histogram for lower performing players assembles the
salaries histogramm for all players but has a less fatter tail. The salaries of 
players with the best performance reach along the range from just below 1 
million USD up to more than 30 million USD: there are still a high number of
best perfoming players earning comparably low salaries while only few reach a 
salary above 30 million USD.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_11}

summary(subset(nba_stats_salar, PER <= 20, select=salary_infl))
summary(subset(nba_stats_salar, PER > 20, select=salary_infl))

```

We can see that ranges of salaries for players with PER above 20 and below or at
20 differ a lot. Also the mean salaries of players with PER above 20 amounts to 
approx. 15.9 million USD which is much higher than the mean salary of players 
with PER below 10 which is only about 5.1 million USD.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots_12}
# Players staying or leaving current team
table(nba_stats_salar$stayed)

```

Overall, over the past ten years 984 players changed their teams during or
after the season, 1270 stayed with the same team as in the previous season. 
I wonder if changing teams is linked to player's salary.

# Univariate Analysis

### What is the structure of your dataset?

The dataset contains data on the past 10 NBA seasons including different 
performance indicators and players data. Overall, there are 2254 ovservations. 
The dataset is structured in such a way that performance statistics of a given 
season corresponds to a player's salary in the next season since salaries in the 
NBA are determined before the season start. The mean salary (adjusted for
inflation is 6 million USD and the median salary is 3.5 million USD. Mean 
players age is 26 years and mean career length among active players over the 
last ten years is about 4.8 years. Performance statistics such points from the
field, free throws, blocks, steals, assists, turnovers, rebounds all have long 
right tail resulting in a small number of players outscoring the average. 

### What is/are the main feature(s) of interest in your dataset?

The main features in the data set are players salaries along with different
performance indicators. Here salary level differs dramatically starting from 
less than one million USD up to over 30 millions USD. I would like to determine 
how players performance is linked to their salary level.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Other features which may be linked to the salary level are player's age and
experience in the NBA as well as whether the player stays with the same team as 
in the previous season.

### Did you create any new variables from existing variables in the dataset?

I created variables indicating per game performance since in the starting 
datasets only overall performance of the player for a team in a given season was
given along with the number of games. I also created a variable named 'stayed' 
indicating whether the player stayed with the same team as in the previous
season.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

I adjusted the data by matching salaries data with games statistics and filtered 
the data so that only observation with more than 5 games per season are 
considered in order to remove outliers. Furthermore, I adjusted salary data for 
inflation since the players performance indicators are comparable over the years
(at least given that the rules of the game are not changed significantly which 
is the case for the time frame being explored).

# Bivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_1}
# correlation matrix
nba_stats_salar_c <- nba_stats_salar %>% 
  dplyr::select('salary_infl', 'Age', 'career_year', 'PPG', 'MPG', 'FTPG', 
                'ASTPG', 'STLPG', 'TRBPG', 'BLKPG', 'TOVPG', 'PFG', 'eFG.',
                'TS.', 'PER')

cor(nba_stats_salar_c)

```

Salary is most correlated with points per game, minutes played, total rebounds
per game and PER. Also minutes played per game correlated strongly with many
'per game' - indicators which makes sense since players have more chances to 
score, assist etc. if they play more. At the same time number of blocks per game 
and shooting accuracy represented by effective field goal and true shooting
percentage are not strongly related to the number of minutes played. Number of
personal fouls per game is correlated most with rebounds which makes also sense 
since rebounding fouls are common. Turnovers are also strongly correlated with
points per game which is a bit surprising to me and I don't have a good 
explanation for that. It might be that players having the ball most of the time
make many points and at the same time have a higher chance of loosing the ball.
Age and career year show only poor correlations with salary as well as with 
players perfomance indicators, so experience surely helps in many aspects of the
game but we cannot yet see it from the variables under our exploration here.
Lets have a closer look at some per game indicators using a plot matrix.


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=10, Bivariate_Plots_2}
# ggpairs matrix
nba_stats_salar_f <- nba_stats_salar %>% 
  dplyr::select('salary_infl', 
                'PPG', 'MPG', 'FTPG', 'ASTPG', 'STLPG', 
                'TRBPG', 'BLKPG')

ggpairs(nba_stats_salar_f, 
        lower = list(continuous = wrap("points", shape = I('.'))),
        upper = list(combo = wrap("points", outlier.shape = I('.'))))

```

I tried to build a linear regression model from some of the variables for
predicting salaries which will be introduced in the multvariate analysis 
section. I didn't use variables which obviously influence each other, e.g. I
didn't use the variable minutes per game (MPG), since it influences other 
variables such as points per game (PPG) or 2 points per game (X2PPG). This is 
also the reason why I didn't use player efficiency rating (PER) since it is 
calculated using other variables I used for my modell, although it has a high
correlation with salary as can be seen from the correlation matrix above.  

Since the plots presented above just use a linear scale I wonder how the
relationship between salary and performance indicators would look like when
ajdusting the scale using log or square root function, since as we've seen in
the section on univariate plots some variables are better explored by adjusting
the scale.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=5, fig.height=10, Bivariate_Plots_3}
# Relationships between salary and performance statistics
# Salary and 2 points goals per game 
s_x2ppg <- ggplot(aes(x = X2PPG, y = salary_infl), data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(trans = sqrt_trans(), 
                     limits = c(quantile(nba_stats_salar$X2PPG, 0.01), 
                             quantile(nba_stats_salar$X2PPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), 
                     limits = c(quantile(nba_stats_salar$salary_infl, 0.01), 
                             quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. 2 points goals per game sqrt') +
  geom_smooth(method = 'lm')

# Salary and 3 points goals per game
s_x3ppg <- ggplot(aes(x = X3PPG, y = salary_infl), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + scale_x_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$X2PPG, 0.01), 
                             quantile(nba_stats_salar$X2PPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                             quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. 3 points goals per game sqrt') +
  geom_smooth(method = 'lm')

# Salary and Free throws goals per game
s_ftpg <- ggplot(aes(x = FTPG, y = salary_infl), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + 
  scale_x_continuous(trans = log_trans()) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                             quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. free throws per game log') +
  geom_smooth(method = 'lm')

grid.arrange(s_x2ppg, s_x3ppg, s_ftpg)

```

From the scoring indicators, 2 point goals and free point goals are mostly 
related to the salary. 3 point goals are not strongly related to the salary:
there are many players with relatively high salary and relatively low 3 
points scoring. Remember, from the histogram plot of 3 point goals we learned 
that there are many players who didn't score a single 3 pointer per game.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=20/3, Bivariate_Plots_4}
# Salary and Assists per game
s_astpg <- ggplot(aes(x = ASTPG, y = salary_infl), data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(trans = log_trans(), limits = 
                                     c(quantile(nba_stats_salar$ASTPG, 0.01), 
                                       quantile(nba_stats_salar$ASTPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Assists per game log') +
  geom_smooth(method = 'lm')

# Salary and Steals per game
s_stlpg <- ggplot(aes(x = STLPG, y = salary_infl), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + scale_x_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$STLPG, 0.01), 
                         quantile(nba_stats_salar$STLPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Steals per game sqrt') +
  geom_smooth(method = 'lm')

# Salary and Total rebounds per game
s_trbpg <- ggplot(aes(x = TRBPG, y = salary_infl), data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(trans = log_trans(), limits = 
                       c(quantile(nba_stats_salar$TRBPG, 0.01), 
                         quantile(nba_stats_salar$TRBPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Total rebounds per game sqrt') +
  geom_smooth(method = 'lm')

# Salary and Blocks per game
s_blkpg <- ggplot(aes(x = BLKPG, y = salary_infl), data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(trans = log_trans()) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Blocks per game log') +
  geom_smooth(method = 'lm')

grid.arrange(s_astpg, s_stlpg, s_trbpg, s_blkpg)

```

Among other performance indicators total rebounds have the strongest 
relationship with the salary.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=5, fig.height=10, Bivariate_Plots_5}
# Salary and Turnovers per game
s_tovpg <- ggplot(aes(x = TOVPG, y = salary_infl), data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(trans = log_trans(), limits = 
                       c(quantile(nba_stats_salar$TOVPG, 0.01), 
                         quantile(nba_stats_salar$TOVPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Total rebounds per game log') +
  geom_smooth(method = 'lm')
# Salary and Personal fouls per game
s_pfg <- ggplot(aes(x = PFG, y = salary_infl), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + scale_x_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$PFG, 0.01), 
                         quantile(nba_stats_salar$PFG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Personal fouls per game sqrt') +
  geom_smooth(method = 'lm')

grid.arrange(s_tovpg, s_pfg)

```

Interestingly, negative performance represented by the number of turnovers per 
game and personal fouls per game also positively related to the salary.


```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=5, fig.height=10, Bivariate_Plots_6}
# Salary and Advanced statistic
# Salary and Effective field goal percentage
s_efg <- ggplot(aes(x = eFG., y = salary_infl), data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(limits = 
                       c(quantile(nba_stats_salar$eFG., 0.01), 
                         quantile(nba_stats_salar$eFG., 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Effective field goal percentage') +
  geom_smooth(method = 'lm')

# Salary and True shooting percentage
s_ts <- ggplot(aes(x = TS., y = salary_infl), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + scale_x_continuous(limits = 
                       c(quantile(nba_stats_salar$TS., 0.01), 
                         quantile(nba_stats_salar$TS., 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. True Shooting') +
  geom_smooth(method = 'lm')

grid.arrange(s_efg, s_ts)

```

Plotting the shooting accuracy variables effective field goal percentage and 
true shooting reveals that there are very few players whos shooting accuracy is 
low but salary is high (the upper left corners of the both plots have only few 
points). At the same time there are plenty players whos shooting accuracy is 
relatively high but the salary is low.

Next, let us look at categorical variables such as position and staying with the
team and salary distribution across them. Additionally let's look at salary vs. 
age and career year using box plots which might deliver a better view than 
points plot.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_7}
# Box plot stayed
boxplot(salary_infl ~ stayed , data = nba_stats_salar, 
        main = 'Salary and staying with the same team', 
        xlab = 'Stayed', ylab = 'Salary')

```

It looks like staying with the same team is connected to a bigger salary 
measured by mean salary and 3rd quartile where both are higher for those who 
stay. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_8}
# Box plot career year
boxplot(salary_infl ~ career_year , data = nba_stats_salar, 
        main = 'Salary by career year', 
        xlab = 'career year', ylab = 'Salary')

```

Also in the first three career years salaries are smaller than in later 
years. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_9}
# Salary box plots
# Box plot position
boxplot(salary_infl ~ Pos , data = nba_stats_salar, main = 'Salary by position', 
        xlab = 'Position', ylab = 'PPG')
```

Differences in salaries between positions also exist. Now let us look into the 
position variable in more detail. Who scores most points? 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_10}
# box plots by position
boxplot(PPG ~ Pos , data = nba_stats_salar, main = 'Points by position', 
        xlab = 'Position', ylab = 'PPG')

```

It looks like point guards and shooting guards score most on average, which 
makes sense, although the difference to other positions is not dramatically 
large. Who does most rebounding?

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_11}

boxplot(TRBPG ~ Pos , data = nba_stats_salar, 
        main = 'Total rebounds by position', xlab = 'Position', 
        ylab = 'Total rebounds per game')

```

Do different positions require different physical characterics like 
height and weight?

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_12}
boxplot(height ~ Pos , data = nba_stats_salar, main = 'Height by position', 
        xlab = 'Position', 
        ylab = 'Height')

boxplot(weight ~ Pos , data = nba_stats_salar, main = 'Weight by position', 
        xlab = 'Position', 
        ylab = 'Weight')

```

Big guys like center and power forwards rebound most which is also no
surpise. Finally, do experienced players score more or draw more fouls?

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_13}
# plots points and free throws per game by career year
boxplot(PPG ~ career_year, data = nba_stats_salar, 
        main = 'Points per game by career year', xlab = 'career year', 
        ylab = 'PPG')

boxplot(FTPG ~ career_year, data = nba_stats_salar, 
        main = 'Free throws by career year', xlab = 'career year', 
        ylab = 'Free throws per game')


```

Maybe the reason for younger players to score less is having less time to 
score?

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_14}
# minutes per game by career year
boxplot(MPG ~ career_year, data = nba_stats_salar, 
        main = 'Minutes played by career year', xlab = 'career year', 
        ylab = 'Minutes played per game')
```

To see if it's true let us proof whether this difference still exists when
points per minute instead of points per game and plot it by career year.

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Plots_15}
# Create and plot new variable minutes per game by career year
nba_stats_salar$points_per_min <- with(nba_stats_salar, PPG/MPG) 

boxplot(points_per_min ~ career_year, data = nba_stats_salar, 
        main = 'Points per min by career year', xlab = 'career year', 
        ylab = 'Points per min')

```

Here we saw that players in their first and second career years seem to have a 
more cautious game drawing less fouls and thus having less free throws per game
as well as less points per game. One part of it can be explained by the fact 
that younger players spent less minutes on the field indicated by minutes per 
game by career year vs. minutes plot. Plotting points per minute for different 
career years reveals that the difference in points in first two years is not 
that big although it exists. Finally, for players who have passed their 10th
career year it is harder to make average judgements since there are fewer data
points for those as we saw by plotting the career year variable in the 
univariate plots section above.

I wonder how performance indicators and their relation to the salary variable
explored above for all players look differently for different positions. Let us 
have a closer look at it later in the multivariate analysis chapter.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

I observed some interesting relationships between salary level and game 
performance indicators. In most cases relationships observed are not linear. 
Relationships between salary and players experience expressed by age and career 
year varibale were not as strong as I expected.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

To get a better understanding about the data I looked at how two important 
variables, points per game and number of free throws, are different for 
different career years by creating box plots. Both plots revealed that in the
beginning of the career players have lower number of overall points and free 
throws per game. Then I created a plot of number of minutes played for different
career years which revealed to me a similar pattern: players with less
experience in the NBA spent less time in the game which may explain their lower
scoring. Plotting points per minute reveals that the difference in scoring for
younger players is not that big although it still exists.

### What was the strongest relationship you found?

The strongest relationship observed was between salary and points per game with 
the correlation of approximately 0.69. Number of free throws per game and total
rebounds per game had second and third largest correlations with salary from all
the variables explored so far.  

# Multivariate Plots Section

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_1}
# density of salary by position and stayed variable plots
ggplot(aes(salary, color = Pos), data = nba_stats_salar) +
  geom_density() + scale_x_continuous(breaks = c(5, 10, 15, 20, 25, 30))
```

There are less center players who earn under 5 million USD than other players. 
There are more shooting guards than other players who earn under 2.5 million 
USD and there are less shooting guard players who earn between 12.5 and 20 
million than other positions.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_2}

ggplot(aes(salary, color = stayed), data = nba_stats_salar) +
  geom_density() + scale_x_continuous(breaks = c(5, 10, 15, 20, 25, 30))

```

Among those who earn above 2.5 million USD there are less players who change 
teams than there are players who stay.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6.5, fig.height=15, Multivariate_Plots_3}
# Salary and 2 Points per game 
s_x2ppg_p <- ggplot(aes(x = X2PPG, y = salary_infl, color = stayed), 
       data = nba_stats_salar) + 
  geom_jitter(alpha = 0.25) + 
  scale_x_continuous(limits = c(quantile(nba_stats_salar$X2PPG, 0.01), 
                             quantile(nba_stats_salar$X2PPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), 
                     limits = c(quantile(nba_stats_salar$salary_infl, 0.01),
                             quantile(nba_stats_salar$salary_infl, 0.99))) +
  ggtitle('Salary sqrt vs. 2 Points per game') + geom_smooth(method = 'lm') +
  facet_wrap(~Pos) + 
  theme(legend.position = c(0.88,0.12))

# Salary and 3 Points per game 
s_x3ppg_p <- ggplot(aes(x = X3PPG, y = salary_infl, color = stayed), 
       data = nba_stats_salar) + 
  geom_jitter(alpha = 0.25) + 
  scale_x_continuous(limits = c(quantile(nba_stats_salar$X3PPG, 0.01), 
                             quantile(nba_stats_salar$X3PPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), 
                     limits = c(quantile(nba_stats_salar$salary_infl, 0.01),
                             quantile(nba_stats_salar$salary_infl, 0.99))) +
  ggtitle('Salary sqrt vs. 3 Points per game') + geom_smooth(method = 'lm') +
  facet_wrap(~Pos)  + 
  theme(legend.position = c(0.88,0.12))

# Salary and Free throws per game 
s_fppg_p <- ggplot(aes(x = FTPG, y = salary_infl, color = stayed), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + 
  scale_x_continuous() + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                             quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. free throws per game') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos) + 
  theme(legend.position = c(0.88,0.12))

grid.arrange(s_x2ppg_p, s_x3ppg_p, s_fppg_p, ncol=1)

```

From the plots of the scoring indicators variables by position we can conclude 
that relationships between 2 point goals and free throws definetly there
independet from players positions. As for the 3 point goals variable where we 
saw before that there is no such strong relationship, we now know this is mainly
because of center players. For point guards the number of 3 point goals is
somewhat related to the salary as shown in the plot.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6.5, fig.height=20, Multivariate_Plots_4}
# Salary and Total rebounds per game
s_trbpg_p <- ggplot(aes(x = TRBPG, y = salary_infl, color = stayed), 
       data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(limits = 
                       c(quantile(nba_stats_salar$TRBPG, 0.01), 
                         quantile(nba_stats_salar$TRBPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Total rebounds per game') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos) + 
  theme(legend.position = c(0.88,0.12))

# Salary and Assists per game
s_astpg_p <- ggplot(aes(x = ASTPG, y = salary_infl, color = stayed), 
       data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(limits = 
                                     c(quantile(nba_stats_salar$ASTPG, 0.01), 
                                       quantile(nba_stats_salar$ASTPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Assists per game') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos) + 
  theme(legend.position = c(0.88,0.12))

# Salary and Steals per game
s_stlpg_p <- ggplot(aes(x = STLPG, y = salary_infl, color = stayed), 
       data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + scale_x_continuous(limits = 
                       c(quantile(nba_stats_salar$STLPG, 0.01), 
                         quantile(nba_stats_salar$STLPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Steals per game') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos) + 
  theme(legend.position = c(0.88,0.12))

# Salary and blocks per game
s_blkpg_p <- ggplot(aes(x = BLKPG, y = salary_infl, color = stayed), 
       data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(limits = 
                       c(quantile(nba_stats_salar$BLKPG, 0.01), 
                         quantile(nba_stats_salar$BLKPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Total blocks per game') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos) + 
  theme(legend.position = c(0.88,0.12))

grid.arrange(s_trbpg_p, s_astpg_p, s_stlpg_p, s_blkpg_p, ncol=1)

```

Now we see that the variables rebounds per game and blocks per game are those
which are related to center players salary. At the same time blocks are not 
important for shooting guards. Even if the relationship exists for multiple 
players positions such as in case of assists for point guards, shooting guards 
and small forwards, the relationships is differently important for different
positions.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6.5, fig.height=10, Multivariate_Plots_5}
# Salary and Turnovers per game
s_tovpg_p <- ggplot(aes(x = TOVPG, y = salary_infl, color = stayed), 
       data = nba_stats_salar) + 
 geom_jitter(alpha = 0.25) + scale_x_continuous(trans = log_trans(), limits = 
                       c(quantile(nba_stats_salar$TOVPG, 0.01), 
                         quantile(nba_stats_salar$TOVPG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Total turnovers per game log') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos)  + 
  theme(legend.position = c(0.88,0.12))

# Salary and Personal fouls per game
s_pfg_p <- ggplot(aes(x = PFG, y = salary_infl, color = stayed), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + scale_x_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$PFG, 0.01), 
                         quantile(nba_stats_salar$PFG, 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. Personal fouls per game sqrt') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos)  + 
  theme(legend.position = c(0.88,0.12))

grid.arrange(s_tovpg_p, s_pfg_p, ncol=1)

```

Again, with negative performance indicators, personal fouls per game and 
turnovers per game across positions we see the same positive relationship with 
the salary as for all players together.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=6.5, fig.height=5, Multivariate_Plots_6}
# Salary and True shooting percentage
ggplot(aes(x = TS., y = salary_infl, color = stayed), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + scale_x_continuous(limits = 
                       c(quantile(nba_stats_salar$TS., 0.01), 
                         quantile(nba_stats_salar$TS., 0.99))) + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                         quantile(nba_stats_salar$salary_infl, 0.99))) + 
  ggtitle('Salary sqrt vs. True Shooting') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos) + 
  theme(legend.position = c(0.88,0.12))

```
Lastly, for true shooting we see almost no relationship with salary for center 
players and some relationship for point guards and shooting guards.

Plots of salary vs. game performance indicators by position divided into two
groups (those who stayed and those who left) reveal different picture for 
different positions and stayed / left group. Therefore not all indicators are 
best suited to be used widely across different positions to evaluate players 
performance. In many cases in the above plots the smoothing line for those who 
change teams is below for those who stay which means that the variable stayed 
could be also used for the salaries forecasting considerations.

I tried to build a linear regression for predicting salary based on average per
game performance indicators described above. I also included the categorical
variable 'stayed' into the modell as well as career year variable. Let us look 
at how the modell works for all players as well as for different positions 
separately by subseting the data. Firstly, I subset the data by setting game 
performance indicators which will be used in the regression by setting them
above zero since some variables will be adjusted using the log function.

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_7}
# first subset the data setting variable values above zero, since some of them 
# will be transformed using log function
nba_stats_salar_non_zero <- nba_stats_salar %>% 
  filter(FTPG > 0 & X2PPG > 0 & X3PPG > 0 & TRBPG > 0 & ASTPG > 0 & STLPG > 0 &
           TOVPG > 0 & PFG > 0 & BLKPG > 0 & TS. > 0)

dim(nba_stats_salar_non_zero)
summary(nba_stats_salar_non_zero)

```

##### Modell for all players positions

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_8}
# all players
m1 <- lm(I(sqrt(salary_infl)) ~ I(career_year), 
         data = nba_stats_salar_non_zero)
m2 <- update(m1, ~ . + stayed) 
m3 <- update(m2, ~ . + FTPG) 
m4 <- update(m3, ~ . + X2PPG)
m5 <- update(m4, ~ . + X3PPG)
m6 <- update(m5, ~ . + TRBPG)
m7 <- update(m6, ~ . + ASTPG)
m8 <- update(m7, ~ . + STLPG)
m9 <- update(m8, ~ . + log(TOVPG))
m10 <- update(m9, ~ . + sqrt(PFG))
m11 <- update(m10, ~ . + BLKPG)
m12 <- update(m11, ~ . + TS.)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)

```

##### Modell for Center

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_9}
# modell Center
m1 <- lm(I(sqrt(salary_infl)) ~ I(career_year), 
         data = subset(nba_stats_salar_non_zero, Pos == 'C'))
m2 <- update(m1, ~ . + stayed) 
m3 <- update(m2, ~ . + FTPG) 
m4 <- update(m3, ~ . + X2PPG)
m5 <- update(m4, ~ . + X3PPG)
m6 <- update(m5, ~ . + TRBPG)
m7 <- update(m6, ~ . + ASTPG)
m8 <- update(m7, ~ . + STLPG)
m9 <- update(m8, ~ . + log(TOVPG))
m10 <- update(m9, ~ . + sqrt(PFG))
m11 <- update(m10, ~ . + BLKPG)
m12 <- update(m11, ~ . + TS.)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)

```

##### Modell for Point Guard

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_10}
# modell Point Guard
m1 <- lm(I(sqrt(salary_infl)) ~ I(career_year), 
         data = subset(nba_stats_salar_non_zero, Pos == 'PG'))
m2 <- update(m1, ~ . + stayed) 
m3 <- update(m2, ~ . + FTPG) 
m4 <- update(m3, ~ . + X2PPG)
m5 <- update(m4, ~ . + X3PPG)
m6 <- update(m5, ~ . + TRBPG)
m7 <- update(m6, ~ . + ASTPG)
m8 <- update(m7, ~ . + STLPG)
m9 <- update(m8, ~ . + log(TOVPG))
m10 <- update(m9, ~ . + sqrt(PFG))
m11 <- update(m10, ~ . + BLKPG)
m12 <- update(m11, ~ . + TS.)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)

```

##### Modell for Shooting Guard

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_11}
# modell Shooting Guard
m1 <- lm(I(sqrt(salary_infl)) ~ I(career_year), 
         data = subset(nba_stats_salar_non_zero, Pos == 'SG'))
m2 <- update(m1, ~ . + stayed) 
m3 <- update(m2, ~ . + FTPG) 
m4 <- update(m3, ~ . + X2PPG)
m5 <- update(m4, ~ . + X3PPG)
m6 <- update(m5, ~ . + TRBPG)
m7 <- update(m6, ~ . + ASTPG)
m8 <- update(m7, ~ . + STLPG)
m9 <- update(m8, ~ . + log(TOVPG))
m10 <- update(m9, ~ . + sqrt(PFG))
m11 <- update(m10, ~ . + BLKPG)
m12 <- update(m11, ~ . + TS.)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)

```

##### Modell for Power Forward

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_12}
# modell Power Forward
m1 <- lm(I(sqrt(salary_infl)) ~ I(career_year), 
         data = subset(nba_stats_salar_non_zero, Pos == 'PF'))
m2 <- update(m1, ~ . + stayed) 
m3 <- update(m2, ~ . + FTPG) 
m4 <- update(m3, ~ . + X2PPG)
m5 <- update(m4, ~ . + X3PPG)
m6 <- update(m5, ~ . + TRBPG)
m7 <- update(m6, ~ . + ASTPG)
m8 <- update(m7, ~ . + STLPG)
m9 <- update(m8, ~ . + log(TOVPG))
m10 <- update(m9, ~ . + sqrt(PFG))
m11 <- update(m10, ~ . + BLKPG)
m12 <- update(m11, ~ . + TS.)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)

```

##### Modell for Small Forward

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots_13}
# modell Small Forward
m1 <- lm(I(sqrt(salary_infl)) ~ I(career_year), 
         data = subset(nba_stats_salar_non_zero, Pos == 'SF'))
m2 <- update(m1, ~ . + stayed) 
m3 <- update(m2, ~ . + FTPG) 
m4 <- update(m3, ~ . + X2PPG)
m5 <- update(m4, ~ . + X3PPG)
m6 <- update(m5, ~ . + TRBPG)
m7 <- update(m6, ~ . + ASTPG)
m8 <- update(m7, ~ . + STLPG)
m9 <- update(m8, ~ . + log(TOVPG))
m10 <- update(m9, ~ . + sqrt(PFG))
m11 <- update(m10, ~ . + BLKPG)
m12 <- update(m11, ~ . + TS.)
mtable(m1, m2, m3, m4, m5, m6, m7, m8, m9, m10, m11, m12)

```

The modell explains only just about 59 % of the salary variable 
indicated by R^2 of 0.59. There are seem to be factors explaining the 
salary level other than players performance on the field. The modell works best 
for power forwards and point guards, the respective R^2 of 0.628 and 0.632 
are the largest compared to other positions. The modell for center players has
the lowest R^2 with 0.537. First, by subsetting the data and setting per game 
performance statistic larger than zero, we are left with only 135 center players
out of 438 in the original data frame since many of them might have not scored a
single 2 or 3 points goal for example. Their performance is better measured by
rebounds or blocks, there could be some other indicators of their performance 
not analyzed here. The modell is obviously not optimal, but still is able to 
show that players perfomance in the game is related to their salary level 
although there are obviously some other factors involved such as salary cap and 
negotiation ability which may be connected to players salary.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

As you can see from last plots by position in the multivariate analysis section,
some game performance indicators show a more or less linear relationships with
salary variable, e.g. points per game variable. Some of them show less linear 
and less stronger relationships or no relationship at all e.g. 3 points per 
game for center players. Different variables work differently for different 
players. So I tried to create a linear modell using most of the variables 
explored above and then test it for different players positions.

### Were there any interesting or surprising interactions between features?

Ranges of variables vary across positions. For example the range of assits per 
game for centers is about one and a half times as small as for point guards. 
This makes sense since point guards are those who handle the ball most and 
initiate the offense as opposed to center players. This range difference 
can be also seen for some other variables across different positions.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

Yes, I created a linear modell which tries to explain players salary through 
variables of his game performance. These variables describe shooting perfomance, 
defence performance and so on. Since different positions have different focus
(offence, defence, assists, rebounding etc.) for every position different
variables work differently and have different R^2 outcomes. The modell for all 
players has an R^2 of about 59%, so the choosen game performance indicators and
other modell variables explain more than a half of players salary. Limitations 
of the modell is that it does not take into account other variables
which may be important for the salary. 

------

# Final Plots and Summary

### Plot One

```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_One}
# salary sqrt transformed histogram
ggplot(aes(x = salary_infl), data = nba_stats_salar) + 
  geom_histogram(binwidth = 0.2) + 
  scale_x_continuous(trans = sqrt_trans(), 
                     breaks = c(0.2, 0.5, 1, 2.5, 5, 10, 20, 30)) +
  scale_y_continuous(breaks = c(50, 100, 150, 200, 250)) +
  ggtitle('Salary histogram sqrt transformed') + 
  xlab('Salary (adj. for inflation) in million USD') +
  ylab('Number of players')

```

### Description One

The distribution of NBA players salaries appears to have a bimodal shape on a 
square root scale. The distribution is skewed to the right with only a small
number of players earning over 20 milllion USD. 

### Plot Two
```{r echo=FALSE, message=FALSE, warning=FALSE, Plot_Two}
# Salary box plots
# Box plot position
boxplot(salary_infl ~ Pos , data = nba_stats_salar, 
        main ='Salary by position', 
        xlab = 'Position', ylab = 'Points per game', 
        col = "lightgrey", par(las = 1))

# Box plot career year
boxplot(salary_infl ~ career_year , data = nba_stats_salar, 
        main = 'Salary by career year', 
        xlab = 'career year', 
        ylab = 'Salary (adj. for inflation) in million USD', 
        col = "lightgrey", par(las = 1))

```

### Description Two

Salaries vary across positions and along with the career progress. Center
players and power forwards earn most measured by the interquartile range, but at
the same time salaries of players in other positions show much more outliers at
the top. Salaries of younger players in their first to third year tend to be 
much smaller compared to those with more advanced careers.

### Plot Three
```{r echo=FALSE, warning=FALSE, message=FALSE, warning=FALSE, Plot_Three}
# Salary and Points per game 
ggplot(aes(x = PPG, y = salary_infl, color = stayed), data = nba_stats_salar) + 
geom_jitter(alpha = 0.25) + 
  scale_x_continuous() + 
  scale_y_continuous(trans = sqrt_trans(), limits = 
                       c(quantile(nba_stats_salar$salary_infl, 0.01), 
                             quantile(nba_stats_salar$salary_infl, 0.99)), 
                     breaks = c(1, 5, 10, 20)) + 
  ggtitle('Salary (sqrt) vs. points per game') +
  geom_smooth(method = 'lm') + facet_wrap(~Pos) + 
  xlab('Points per game') + ylab('Salary (adj. for inflation) in million USD') +
  scale_colour_discrete(name  ="Stayed with the team?",
                          breaks=c(TRUE, FALSE),
                          labels=c("Yes", "No")) + 
  theme(legend.text = element_text(size = 12), legend.position = c(0.88,0.12))

```

### Description Three

Players salary (transformed using square root function) and diverse game
performance indicators appear to be related to each other. Here an example of 
the relationship between salary and points per game is shown. Different 
performance metrics showed different relationships to players salary across 
different positions. I used some of the per game performance indicators such as 
2 and 3 point goals, free throws, blocks, assists, rebounds and steals along
with career year and 'stayed' variable to build a linear regression model. The
stayed variable represents the fact whether or not the player changed the team 
within or after the season. The modell explains about 60% of players salary 
(based on all players) and deliver slightly different results across positions 
with the highest R of approx. 0.63 for point guards and power forwards. The 
modell fails to explain the rest of salary since there might be other factors 
influencing the salary ranging from salary cap rules to player's agent's ability
to negotiate.

------

# Reflection

The NBA salaries and performance dataset contains salaries of NBA players over
the last 10 years along with their average game performance metrics and some
other players charateristics such as age, weight and height. I started by
understanding choosen variables in the dataset using univariate plots as well as
how variables are connected to each other using bivariate and multivariate 
plots. I then tried to build a modell explaining NBA salaries through player's
performance and experience.

I went through some struggles trying to understand how variables in the dataset
are related to each other and influenced by each other with an example of the 
position variable influencing one or another performance indicator. Another 
challenge was to understand the structure of different datasets I used as input 
in the beginning to pull together the final dataset used for analysis here. As I 
was expecting the salary is connected to players performance. At the same time
performance indicators vary across positions. It was a surprise to me that 
players who stay with their teams earn more on average than those who change 
teams.

Future work which can be done using the dataset is exploring other variables 
in the dataset, trying to build a better modell which could explain more than 
just 60% of the salary by finding and incorporating new variables.